---
pdf_document: default
author: "Felix Ehis Osawaru"
date: "06/04/2022"
output:
  word_document: default
  html_document: default
  pdf_document: default
title: "Data Vitualisation and Analysis"
---

```{r setup, include=TRUE}
knitr::opts_chunk$set(echo = TRUE)
```

```{r}
library(ggplot2)
```

```{r}
library(plyr)
```

```{r}
library(dplyr)
```

```{r}
library(lattice)
```

```{r}
library(Rmisc)
```

```{r}
plankton <- read.csv("plankton.csv", header=T, stringsAsFactors = T)
```
                                    QUESTION 1
```{r}
summary(plankton)
```
```{r}
str(plankton)
```
```{r}
class(plankton)
```
##Pseudonitzschia.A.Sp
```{r}
summary(plankton$Pseudonitzschia.A.Sp)
```
## Obtaining the statistical value for plankton and Pseudonitzschia.A.Sp
```{r}
mean(plankton$Pseudonitzschia.A.Sp)
median(plankton$Pseudonitzschia.A.Sp)
sd(plankton$Pseudonitzschia.A.Sp)
max(plankton$Pseudonitzschia.A.Sp)
IQR(plankton$Pseudonitzschia.A.Sp)
range(plankton$Pseudonitzschia.A.Sp)
var(plankton$Pseudonitzschia.A.Sp)
```
##Alexandrium.Sp
```{r}
summary(plankton$Alexandrium.Sp)
```
## Obtaining the statistical value for plankton and Alexandrium.Sp
```{r}
mean(plankton$Alexandrium.Sp)
median(plankton$Alexandrium.Sp)
sd(plankton$Alexandrium.Sp)
min(plankton$Alexandrium.Sp)
max(plankton$Alexandrium.Sp)
IQR(plankton$Alexandrium.Sp)
range(plankton$Alexandrium.Sp)
var(plankton$Alexandrium.Sp)
```
## Robgordia.Sp
```{r}
summary(plankton$Robgordia.Sp)
```
## Obtaining the statistical value for plankton and Robgordia.Sp
```{r}
mean(plankton$Robgordia.Sp)
median(plankton$Robgordia.Sp)
sd(plankton$Robgordia.Sp)
min(plankton$Robgordia.Sp)
max(plankton$Robgordia.Sp)
IQR(plankton$Robgordia.Sp)
range(plankton$Robgordia.Sp)
var(plankton$Robgordia.Sp)
```

## water.Temp
```{r}
summary(plankton$Water.Temp)
```
```{r}
mean(plankton$Water.Temp)
median(plankton$Water.Temp)
sd(plankton$Water.Temp)
min(plankton$Water.Temp)
max(plankton$Water.Temp)
IQR(plankton$Water.Temp)
range(plankton$Water.Temp)
var(plankton$Water.Temp)
```
## Region
```{r}
counts <- table(plankton$Region)
counts
```
```{r}
prop.table(counts)
```
##Species
```{r}
counts <- table(plankton$Species)
counts
```

```{r}
prop.table(counts)
```
## Year
```{r}
counts <- table(plankton$year)
counts
```

```{r}
prop.table(counts)
```
## Period
```{r}
counts <- table(plankton$period)
counts
```
```{r}
prop.table(counts)
```
                                 
                                 QUESTION 2
## Using Boxplot to show the distribution of Pseudonitzschia
```{r}
p <- ggplot(plankton, aes(y=Pseudonitzschia.A.Sp))
p <- p + geom_boxplot(fill="lightblue",)
p <- p + labs(x="Pseudonitzschia.A.Sp",
         y="Count",
         title = "Boxplot for Pseudonitzschia.A.Sp")+ 
         theme(plot.title = element_text(hjust = 0.5))
p
```

The distribution is positively skewed, the number of outliers is within the
upper bound of the data, and the median does not divide the box evenly.
That is, the mean, the standard deviation is higher than the median.

## Water.Temp
```{r}
p <- ggplot(plankton, aes(y=Water.Temp))
p <- p + geom_boxplot(fill="lightblue",)
p <- p + labs(x="Water.Temp",
         y="Count",
         title = "Boxplot for Water.Temp")+
         theme(plot.title = element_text(hjust = 0.5))
p
```

The distribution is symmetric because the whiskers and outliers on the left are
almost the same as on the right, and there is an equal amount of data in each 
quadrant. 

                                    QUESTION 3
## Using Univeriate statistics to compare Pseudonitzchia data for 2021 and
## before 2021 
```{r}
Pseudonitzchia.A.sp2021 <- plankton$Pseudonitzschia.A.Sp[plankton$year== 2021]
summary(Pseudonitzchia.A.sp2021)
```
##Obtaining the statistical value for plankton$Pseudonitzschia.A.Sp for 2021
```{r}
sd(Pseudonitzchia.A.sp2021)
range(Pseudonitzchia.A.sp2021)
var(Pseudonitzchia.A.sp2021)
IQR(Pseudonitzchia.A.sp2021)
```

## Pseudonitzchia distribution Before 2021 
```{r}
Pseudonitzchia.A.sp2021 <- plankton$Pseudonitzschia.A.Sp[plankton$year != 2021]
summary(Pseudonitzchia.A.sp2021)
```

##Obtaining the statistical value for plankton$Pseudonitzschia.A.Sp before 2021
```{r}
sd(Pseudonitzchia.A.sp2021)
range(Pseudonitzchia.A.sp2021)
var(Pseudonitzchia.A.sp2021)
IQR(Pseudonitzchia.A.sp2021)
```

Comparing the data from 2021 to those before 2021, it can be seen that the mean, 
median, and standard deviation of 2021 are higher than the data of previous 
years. It suggests that Pseudonitzschia.A.Sp species in 2021 are more dispersed
than in previous years.

                                  QUESTION 4
## Plotting the Histogram of Normal and Skewed distribution
```{r}
p <- ggplot(data =plankton, aes(Water.Temp))
p <- p + geom_histogram(colour="white", fill="lightblue", binwidth =2)
p <- p + labs(x="Water.Temp",
         y="Count",
         title = "Histogram for Water.Temp")+
         theme(plot.title = element_text(hjust = 0.5))
p
```

This is a normal distribution because its bell curve has a peak, and the mean, 
median, and mode are equal.l.

```{r}
p <- ggplot(data =plankton, aes(Pseudonitzschia.A.Sp))
p <- p + geom_histogram( fill="lightblue",bins=20)
p <- p + labs(x="Pseudonitzschia.A.Sp",
         y="Count",
         title = "Histogram for Pseudonitzschia.A.Sp")+ 
         theme(plot.title = element_text(hjust = 0.5))
p
```

Positively skewed distribution because the long tail moves to the right. 
And the value is to the left of the mean. The mean is greater than the median. 
The mean is close to the right side of the distribution and the median is close 
to the left side of the distribution

                                   QUESTION 5
## Barplot to represent Data
```{r}
p <- ggplot(plankton, aes(Species, fill=Species)) 
p <- p+ geom_bar(width = 1)
p <- p+ geom_bar(position = "stack")
p<- p+ labs(x="Species", 
       title ="Bar plot for Species",)+
       theme(plot.title = element_text(hjust = 0.5))
p
```

From the Bar graph, we can say that the Common mussels have the highest count 
in the species and the Razors with the lowest count.

                                QUESTION 6
## plotting the pie chart for water temperature 
```{r}
p <- ggplot(plankton, aes(x="", y= Water.Temp, fill = period))
p <- p + geom_bar(width = 1, stat = "identity")
p <- p + coord_polar ("y", start = 0)
P <- p + theme_void()
p <- p + labs(x="Water Temperature", title = "Pie chart showing water 
              Temperature Distribution")
p
```
The pie chart shows that the water temperature is lower in the first half of 
the year and the highest in the second half of the year.

                                QUESTIION 7
##Plot showing values of Pseudonitzschia.A.Sp against values of Robgordia.S
```{r}
p <- ggplot(plankton, aes(Robgordia.Sp,Pseudonitzschia.A.Sp, colour=Species))
p <- p + geom_point()
p<- p+ labs(x="Robgordia.Sp", y="Pseudonitzschia.A.Sp", title ="Scatter plot of 
            Robgordia.Sp against Pseudonitzschia.A.Sp",) +theme(plot.title =                               element_text(hjust = 0.5))
p
```

The values of Pseudonitzschia and Robgordia have a negative correlation because
the value species Common mussels and pacific oysters are very close to the 
straight line. However, the three Common mussels points represent a negative 
residual,  smaller than the predicted value.
                                
                                QUESTION 8
## Plotting a graph for of Alexandrium.Sp in different regions by species
```{r}
p <- ggplot(plankton, aes(x = Region, y = Alexandrium.Sp, colour = Species))
p <- p + geom_point(position = "jitter", pch = 16, size = 3 )
p <- p + labs(y = "Alexandrium.Sp",
title ="Scatter plot of Alexandrium.Sp in different regions")
p
```

The plot above shows that there is a positive tendency among Alexandrium.Sp the
species increase in different regions as Alexandrium.Sp value increases.
Common mussels have an outlier furthest from the regression line, which is 3000.
Adding jitter to the plot helps separate overlapping points to show the 
descriptive relationship.

                                    Question 9
## Discovering a pair of plankton species which are correlated and a pair which 
```{r}
cov(plankton$Pseudonitzschia.A.Sp,plankton$Robgordia.Sp)
cov(plankton$Pseudonitzschia.A.Sp,plankton$Alexandrium.Sp)
cor(plankton$Pseudonitzschia.A.Sp,plankton$Robgordia.Sp)
cor(plankton$Pseudonitzschia.A.Sp,plankton$Alexandrium.Sp)
```

From the observation, it can be assume that pairs (Pseudonitzschia. A.Sp, 
Robgordia.Sp) has a strong correlation because their value is 0.975273 which is 
close to 1. While the other pair (Pseudonitzschia.A.Sp, Alexandrium.Sp) value 
is 0.06138349 which is not close to 1 and the covariance has a positive number 
hence there is no correlation between these pairs of species
                                  
                                  
                                  QUESTION 10
## Line plot of tempreture against Sample
```{r}
plankton %>% filter(Species%in%c("Common cockles", "Pacific oysters")) %>%
ggplot(aes(x=Sample, y=Water.Temp, colour = Species)) +
geom_point() +
geom_line() +
labs(x = "Sample",
y = "Water.Temp",
title ="Line plot of water temperature against sample index")
```
                                 QUESTION 11
##Producing a linear regression model of Pseudonitzschia.A.Sp on Robgordia.Sp 
##forCommon mussels                                  
```{r}
 lm.output <- lm(formula = Pseudonitzschia.A.Sp ~ Robgordia.Sp, data = plankton)
summary(lm.output)
```

The above statistics show that the R-squared value is 0.9511, which indicates 
the value of Pseudonitzschia.A.Sp and Robgordia are highly correlated, which 
shows a good model. The F-statistic value of 1.464e+04 is very significant for 
the p-value < 2.2e-16
 
## Estimating the value of Pseudonitzschia.A.Sp for the value (1000,2500,4000)
```{r}
newdata <- data.frame(Robgordia.Sp=c(1000,2500,4000))
predict(lm.output,newdata)
lmData <- 
data.frame(residuals = lm.output$residuals, 
                 Pseudonitzschia.A.Sp=plankton$Pseudonitzschia.A.Sp, 
                 Robgordia.Sp= plankton$Robgordia.Sp)
```
## ploting a graph for residuals
```{r}
p <- ggplot(lmData, aes(x=Robgordia.Sp, y = residuals)) 
p <- p + geom_point(size=3, colour="lightblue")
p <- p +  theme_classic() 
p <- p + theme(text = element_text(size = 20))
p
```
## Using the spiro test to check the residual distribution is normal
```{r}
shapiro.test(lmData$residuals)
```

Using the Shapiro-Wilk normality test, the p-value is smaller than 0.05. 
This value represents 0.000000000000000022, which is very close to zero. 
Therefore, the residuals are not considered to be normally distributed. 

```{r}
p <- ggplot(plankton,aes(x = Pseudonitzschia.A.Sp, y= Robgordia.Sp))
p <- p + geom_point() + stat_smooth() 
p<- p+ labs(x="Pseudonitzschia.A.Sp", 
       title ="Scaterplot for values of Pseudonitzschia.A.Sp& Robgordia.Sp ")+
       theme(plot.title = element_text(hjust = 0.5))

p
```

A linear regression model is significant because the residual scatterplots are 
randomly distributed and the graph above shows a linearly increasing relationship
between Pseudonitzschia.A.Sp and Robgordia.Sp. However, the P-value for the 
Shapiro-Wilk normality test is less than 0.05, indicating that the null hypotheses
is rejected, which is contradictory and that is my concern.  

                                   QUESTION 12                                
## Observing if the temperature of the mean is 12 degrees at 99 confidence.
## firstly create a dataframe with columns mean of the temperature of water, 
##month year period.
##creating the dataframe meanWater
```{r}
newdata <- plankton[,c(5,10,11)]
meanWaterTemp<-aggregate(newdata$Water.Temp, by=list(newdata$month, newdata$year), 
FUN=mean, na.rm=TRUE)
colnames(meanWaterTemp) <- c("month","year","meanwatertemp")
View(meanWaterTemp)
summary(meanWaterTemp)

```

```{r}
CI(meanWaterTemp$meanwatertemp, ci=0.99)
```

At the 99% confidence interval, the mean water temperature is 12.13. So, we are 
99% sure that the average temperature is 12 degrees. Without further testing,
95% were confident that the average temperature was not 12.5 degrees.
                                 
                                 QUESTION 13

```{r}
plankton$year.period <- plankton$period
plankton$year.period <- as.numeric(plankton$year.period)
```

```{r}
p <- ggplot(plankton, aes(sample = Water.Temp))
p <- p + stat_qq()
p <- p + stat_qq_line()
p
```
## Shapiro-walk normality test
```{r}
shapiro.test(plankton$Water.Temp)
```

It can be concluded that the dataset passes both tests. Firstly, in the QQ plot, 
the points are closer to a straight line, and the Shapiro-Wilks test shows a 
P-value greater than 0.05. This fact shows that the null hypothesis is true and
the alternative hypothesis is false and there is a correlation between water 
temperature and the time of year. So, the suspicion is justified.
                               
                                QUESTION 14
## using ANOVA test to differentiate species formed in water temperature.

The ANOVA test tells us whether there is a difference between the means. So, to
check for the validity of the Anova test, you apply the Q-Q plot test and the 
Shapiro-Wilk test. If the dot for the Q_Q plot is close to the line and the 
p-value for the Shapiro-Wilk test is higher than 0.05, It is reasonable to 
assume the distribution is normal. However, the ANOVA test's limitation is that
it can only use to investigate a single variable. When distinguishing between 
the means of three or additional categories, it can tell us if at least one pair
of means is significantly different, but not which pair it is.
                                      
                                      QUESTION 15

```{r}
p <- ggplot(data =plankton, aes(month))
p <- p + geom_histogram( fill="lightgreen",binwidth=1)
p <- p + labs(x="month",
             y="Count",
             title = "Histogram for month")+ 
             theme(plot.title = element_text(hjust = 0.5))
p
```

```{r}
p <- ggplot(plankton, aes(y=Region))
p <- p + geom_boxplot(fill="lightgreen",)
p <- p + labs(x="Region",
         y="Count",
         title = "Boxplot for Region")+
         theme(plot.title = element_text(hjust = 0.5))
p
```

From the above graph, we can observe that AGB is the area with the least counts
and SIC is the area with the most counts. While for the other plot, the half of 
the month has the highest count and decreases the count at the start and end of 
the month. Comparing this plot, we can assume that the histogram is negatively 
skewed and that the bar graph increase along each region. The count increases
as data move from one region to another. Although there is an undercount in the
data from one month to another



